A new method for genome-wide marker development and typing holds great promise for primate population genetics

### 2012-12-14 paper/paper.tex
Added abstract and discussion from Luca

\begin{abstract}
\textcolor{red}{\ldots (This is just a bit of the blurb from the email.) Our paper is an introduction to a 2nd generation sequencing technique for typing thousands of genome-wide markers from non-model organisms.  We demonstrate it in five primates and discuss its promise for doing mutli-locus phylogenetics and population genetics in primates. \ldots}
\end{abstract}

@@

[What is mol. primatology? Explain importance of developing markers across taxa in order to study primates at different levels (population, phylogeography, phylogenetics)] During the first decades of molecular primatology, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest. The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011). \textcolor{red}{Due to the bottleneck of marker discovery}, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or few microsatellites (\textcolor{red}{Ting and Sterner, 2012}). Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006). Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).

@@

The following is a summary of the RAD tag library preparation of Etter et al., 2011 (\textcolor{red}{Fig. 1}). The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a). The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site. The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1b). Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1c). To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length. The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1d). It is ligated to the fragments and then the fragments are amplified via PCR (1e). Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence. This ensures that only RAD tags ligated to P1 are able to amplify. After few cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.

@@

We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site. We also removed reads with any quality scores below 10. Using the program Stacks, we clustered all reads into matching sets and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (\textcolor{red}{Stacks ref}). We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).

\textcolor{red}{[Phylogeny?]}

@@

To assess the RAD tag coverage, we mapped human and chimpanzee reads to the human reference genome. Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site. We aligned reads to the human reference genome (\textcolor{red}{hg19, ref}) using BWA with default parameters. We removed reads that were unmapped or that had low mapping quality after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).

\textcolor{red}{Cut this paragraph?} After performing local realignment around indels with GATK (\textcolor{red}{ref}), we identified SNPs and short indels using SAMtools mpileup and BCFtools (\textcolor{red}{ref}). We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location. We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).

To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}). This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).

@@

% 3804227030 bp total (12,983,710 pairs of reads * (142bp + 151bp). This includes non-OWM taxa, remember.
% OWM sequencing reads: 6,125,860 total, half that many pairs
% Of those, 2029586 passed the filtration steps

\_\_\_ million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study. Roughly \_\_\_ million of those reads passed quality control filtration and \_\_\_ could be assigned to a stack. In these stacks our study identified \_\_\_ SNPs among all samples. Information for each individual is summarized in Table 1.

\begin{table}[h]
\caption{Clustered Reads Data}
\begin{center}
	\small
	\begin{tabular}{ p{3cm} || l || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
		\hline
		Taxon & Source & \# Reads & \# Filtered Reads & \# Stacks & Mean Coverage & \# SNPs \\ \hline\hline
		\emph{Microcebus \textcolor{red}{murinus?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
		\emph{Cebus \textcolor{red}{sp.?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
		\emph{Pan troglodytes} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
		\emph{Homo sapiens} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
		\hline
	\end{tabular}
\end{center}
\end{table}

In the human genome, we found \_\_\_ possible cut sites for \emph{PspXI} and \_\_\_ possible sequencing sites (two per cut site, one upstream and one downstream). Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee. For \_\_\_ sites (\_\_\%), both chimpanzee and human had at least one read. When we restrict the analysis to sites with at least three reads, \_\_\_ sites (\_\_\_\%) were covered in human, \_\_\_ sites (\_\_\_\%) were covered in chimpanzee, and \_\_\_ sites (\_\_\_\%) in both.

\begin{table}[h]
\caption{Mapped Reads Data}
\begin{center}
	\small
	\begin{tabular}{ p{3cm} || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
		\hline
		Taxon & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
		\emph{Pan troglodytes} &  &  &  &  &  \\ \hline
		\emph{Homo sapiens} &  &  &  &  &  \\ \hline
		\hline
	\end{tabular}
\end{center}
\end{table}

%\begin{itemize}
%	\item Pairwise comparisons between individuals? Like X orthologous regions shared between Papio and Thero?
%\end{itemize}
%
%%\includegraphics{figs/seq_site_coverage_by_ind}
%
%\begin{itemize}
%	\item SNP info from merged analysis
%	\item SNP Venn diagram?
%	\item Count orthologous SNPs shared between individuals. Pairwise?
%\end{itemize}

@@

\textcolor{red}{[Why it would be good in primates.]}

RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.

\begin{itemize}
	\item Cheap and easy to create libraries. Competent lab can do it in two days. (Ha!)
	\item Multiplexing is great for pop gen studies.
	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
	\item Baird: "different marker densities can be attained by choice of restriction enzyme"
\end{itemize}

Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.

Potential problems with RADseq method including the phylogenetic distance and the need of high quality DNA (at least for now)

ddRADseq

\section{Acknowledgements}
\begin{itemize}
	\item NYU Med School folk
	\item \textcolor{red}{Grants?}
\end{itemize}

>>>>

\begin{abstract}
Over the last two decades primatologists have benefited from the use of numerous molecular markers to study various aspects of primate behavior and evolutionary history. However, most of the studies to date have been based on a single locus (usually mitochondrial DNA) or a few nuclear markers (e.g., microsatellites). Unfortunately, the use of such markers is not only unable to successfully address important questions in primate population genetics and phylogenetics (mainly due to the discordance between gene tree and species tree), but their development is often a time-consuming and expensive task. Today, the advent of next-generation sequencing (NGS) allows researchers to generate large amount of genomic data for non model organisms. However, whole genome sequencing is still cost prohibitive for most primate species. In this paper, we introduce a second generation sequencing technique for typing thousands of genome-wide markers for non-model organisms. Restriction site-Associated DNA Sequencing (RAD-Seq) reduces the complexity of the genome and allow cheap and fast discovery of thousands of markers in many individuals. Here, we describe the principles of this technique and we demonstrate its application in five primates, representing some of the major lineages within the order. Finally, we discuss the promise and possible limitations of such method for doing multi-locus phylogenetics and population genetics in primates.
\end{abstract}

@@

In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research. During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest. The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011). Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or a few microsatellites (Ting and Sterner, 2012). Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006). Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).

@@

The following is a summary of the RAD tag library preparation protocol of Etter et al., 2011 (Fig. 1). The RAD-seq library preparation begins when genomic DNA is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1A). The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site. The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1B). Once the barcode has been added, fragments from multiple individuals can be pooled (1C), and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1D). To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length. The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1E). It is ligated to the fragments and then the fragments are amplified via PCR (1F). Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence. This ensures that only RAD tags ligated to P1 are able to amplify. After few cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.

@@

As input for the clustering analysis, we combined the first read of the paired-end run and the single-end run reads. We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site. We also removed reads with any quality scores below 10. Using the program Stacks, we clustered all reads into sets that differed by no more than two basepairs and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (Catchen et al., 2011). We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).

@@

To assess the RAD tag coverage, we mapped human and chimpanzee reads to the highest quality primate reference genome, that of humans. Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site. We aligned reads to the human reference genome (GRCh37/hg19, Human Genome Consortium 2001) using BWA with default parameters (Li and Durbin, 2009). We separately mapped the single-end and paired-end data and then combined the resultant files after alignment. We removed reads that were unmapped or that had low mapping quality using Picard (http://picard.sourceforge.net) and BamTools (Barnett et al., 2011).

After performing local realignment around indels with GATK (DePristo et al., 2011), we identified SNPs and short indels using SAMtools mpileup and BCFtools (Li et al., 2009). We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location. We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).

To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 basepairs upstream and downstream. (Meyer et al., 2012). This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (Quinlan and Hall 2010).

@@

12.3 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study. Roughly 9.1 million of those reads passed quality control filtration and were clustered into stacks. By comparing these stacks and including only SNPs that were present in multiple individuals, our study identified 7,910 SNPs among all samples. Information for each individual is summarized in Table 1.

\begin{table}[h]
\caption{Clustered Reads Data}
\begin{center}
	\small
	\begin{tabular}{ p{3cm} || l || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
		\hline
		Taxon                       & \# Reads  & \# Filtered Reads & \# Ind. Stacks & Mean Coverage (SD)  & \# Shared SNPs \\ \hline\hline
		\emph{Microcebus sp.}       & 2,830,832 & 2,025,103       & 248,324         &  7.66 (20.71)  & 13    \\ \hline
		\emph{Cebus sp.}            & 1,946,096 & 1,427,413       & 107,829         & 12.64 (139.08) & 56    \\ \hline
		\emph{Theropithecus gelada} & 1,918,425 & 1,392,709       & 136,657         &  9.70 (17.34)  & 212   \\ \hline
		\emph{Pan troglodytes}      & 2,616,062 & 1,910,560       & 157,775         & 11.50 (42.27)  & 5,886 \\ \hline
		\emph{Homo sapiens}         & 3,032,823 & 2,374,733       & 131,544         & 17.37 (25.30)  & 5,786 \\ \hline
	\end{tabular}
\end{center}
\end{table}

In the human genome, we found 58,172 possible cut sites for \emph{PspXI} and 116,344 possible sequencing sites (two per cut site, one upstream and one downstream). Of those possible location, 111,686 sites (96.00\%) had at least one mapped read present in human and 91,646 sites (78.77\%) in chimpanzee. For 90,022 sites (77.38\%), both chimpanzee and human had at least one read. When we restrict the analysis to sites with at least three reads, 109,098 sites (93.77\%) had sequences in human, 89,628 sites (77.04\%) in chimpanzee, and 86,604 sites (74.44\%) in both. From this data, we found 9,275 SNPs relative to the human reference genome that were present in both chimpanzee and human datasets.

\begin{table}[h]
\caption{Mapped Reads Data}
\begin{center}
	\small
	\begin{tabular}{ p{3cm} || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
		\hline
		Taxon                  & \# Reads  & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
		\emph{Pan troglodytes} & 3,917,046 & 2,826,643           &  91,646              &  89,628               & 309,703 \\ \hline
		\emph{Homo sapiens}    & 4,542,978 & 3,784,192           & 111,686              & 109,098               &  35,651 \\ \hline
	\end{tabular}
\end{center}
\end{table}

@@

RAD-Seq represents a fast, cheap, and efficient method for SNP discovery also in species for which no genome reference is available. These characteristics makes RAD-seq an extremely promising techniques for researchers interested in primate population genomics and phylogentics. There are several advantages in using RAD-seq over other molecular techniques. First, this methodology is quite cheap and requires little labwork. The development of a library can be completed in only two days of labwork and all the different steps can be easily performed in a standard molecular lab. Also, the possibility to multiplex several individuals in the same Illumina run either using standard barcodes or a custom combinatorial indexing method (see Peterson et al., 2012 for details) allow researchers to reduce the number of sequencing runs decreasing the costs even further (Peterson et al., 2012; Davey et al., 2011; McCormack, et al. 2012). Second, RAD-seq represents a great improvement in discovering molecular markers to be used in population genetics and phylogenetics. Previous studies to date have been based on single locus (mainly mitochondrial DNA) or a few tens of loci (microsatellite for population genetics or nuclear loci for phylogenetics). RAD-seq techniques can easily produce thousands of independent SNPs in a single run increasing 100-1000x the amount of data available to researchers. Finally, RAD sequencing can produce a large amount of orthologous SNP data that can be employed on a wide range of studies, including population genomics and demographics (e.g., effective population size estimates, bottlenecks, etc.), gene flow and hybridization between closely related species, species boundaries, phylogeography, and phylogeny especially at the intrageneric level (Hohenlohe et al., 2010, 2011; Emerson et al., 2010; Keller et al., 2012; Wagner et al., 2012; Rubin et al., 2012). In summary, we believe that the use of RAD-seq technology can provide extremely valuable information to study recent radiation within primates and to address some major open questions in primatology.

Despite the great potential of the application of RAD sequencing in primatology, we also need to point out some possible limitations of this technique. Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need of high quality samples in order to build the library. In this study we employed high quality DNA, extracted from tissue or blood. However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces. \textcolor{red}{[I'd like to cut this, since I don't want to hint that we're working on techniques to get RAD tags out of feces] Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed. In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples. Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010). It is likely that similar methods can be developed soon allowing primatologists to employ RAD-seq also on low quality DNA samples, such as feces or museum specimens.}

Another possible limitation of the RAD-seq approach is the time scale of its application. RAD-seq data in fact might be not suitable for comparing taxa very distantly related (Rubin et al., 2012). In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a major determinant of the success of the RAD method (Rubin et al., 2012). The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012). This correlation between accuracy and divergence time either reduces the number of orthologus loci available for phylogenetic reconstruction or increase the amount of missing data; both scenarios can affect phylogenetic performance reducing the support values in many nodes or supporting different topologies. However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya. This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).

This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address several questions in primatology, ranging from population genomics to phylogenetics. Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available. Further developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.

\section{Acknowledgements}

\textcolor{red}{Leakey Foundation Grant} The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
